Multi-state modelling with R: the {\tt msm} package \vskip 0.2in \large{Version <<echo=FALSE,results=tex>>= cat(version) @ \vskip 0.1in <<echo=FALSE,results=tex>>= cat(format(Sys.time(), " @ }

### 2014-01-01 pkg/inst/doc/msm-manual.Rnw
Build test again, up to and including first fev fit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@41 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

>>>>

\subsection{Extractor functions}

As well as the functions described in section \ref{sec:extractor} for
extracting useful information from fitted models, there are a number
of extractor functions specific to models with misclassification.

\begin{description}
\item[Misclassification matrix] The function \Rfunction{ematrix.msm}
gives the estimated misclassification probability matrix at the given
covariate values.  For illustration, the fitted misclassification
probabilities for men and women in model 6 are given by

<<>>=
ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
@
The confidence intervals for the estimates for women are higher, since
there are only 87 women in this set of 622 patients.

\item[Odds ratios for misclassification] The function
  \Rfunction{odds.msm} would give the estimated odds ratios corresponding to
  each covariate effect on the misclassification probabilities.

\begin{Scode}
odds.msm(cavmiscsex.msm)
\end{Scode}

\item[Observed and expected prevalences]

  The function \Rfunction{prevalence.msm} is intended to assess the
  goodness of fit of the hidden Markov model for the \emph{observed}
  states to the data.  Tables of observed prevalences of observed
  states are calculated as described in
  section \ref{sec:model-assessment}, by assuming that observed states
  are retained between observation times.

  The expected numbers of individuals in each observed state are
  calculated similarly. Suppose the process begins at a common time
  for all individuals, and at this time, the probability of occupying
  \emph{true} state $r$ is $f_r$.  Then given $n(t)$ individuals under
  observation at time $t$, the expected number of individuals in true
  state $r$ at time $t$ is the $r$th element of the vector $n(t) f
  P(t)$. Thus the expected number of individuals in \emph{observed}
  state $r$ is the $r$th element of the vector $n(t) f P(t) E$, where
  $E$ is the misclassification probability matrix.

  The expected prevalences (not shown) for this example are similar to
  those forecasted by the model without misclassification, with
  underestimates of the rates of death from 8 years onwards.  To
  improve this model's long-term prediction ability, it is probably
  necessary to account for the natural increase in the hazard of death
  from any cause as people become older.

\item[Goodness-of-fit test] The Pearson-type goodness-of-fit test is
  performed, as in Section \ref{sec:pearson}.  The table of deviances
  indicates that there are more 1-3 and 1-4 transitions than expected
  in short intervals, and fewer in long intervals.  This may indicate
  some time-dependence in the transition rates. Indeed, Titman
  \cite{titman:phd} found that a model with piecewise-constant
  transition intensities gave a greatly improved fit to these data.
<<>>=
pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
@

\subsection{Recreating the path through underlying states}

In speech recognition and signal processing, {\em decoding} is the
procedure of determining the underlying states that are most likely to
have given rise to the observations. The most common method of
reconstructing the most likely state path is the {\em Viterbi}
algorithm. Originally proposed by Viterbi \cite{viterbi}, it is also
described by Durbin \etal \cite{biolog:seq} and Macdonald and
Zucchini \cite{macdonald:zucchini} for discrete-time hidden Markov
chains.  For continuous-time models it proceeds as follows. Suppose
that a hidden Markov model has been fitted and a Markov transition
matrix $P(t)$ and misclassification matrix $E$ are known.  Let
$v_k(t_i)$ be the probability of the most probable path ending in state
$k$ at time $t_i$.
\begin{enumerate}
\item Estimate $v_k(t_1)$ using known or estimated initial-state
  occupation probabilities.
\item For $i = 1 \ldots N$, calculate $v_l(t_i) = e_{l,O_{t_i}} \max_k
  v_k(t_{i-1}) P_{kl}(t_{i} - t_{i-1})$. Let $K_i(l)$ be the
  maximising value of $k$.
\item At the final time point $t_N$, the most likely underlying
  state $S^*_N$ is the value of $k$ which maximises $v_k(t_N)$.
\item Retrace back through the time points, setting $S^*_{i-1} =
  K_i(S^*_i)$.
\end{enumerate}

The computations should be done in log space to prevent underflow.
The \Rpackage{msm} package provides the function \Rfunction{viterbi.msm} to
implement this method.  For example, the following is an extract from
a result of calling \Rfunction{viterbi.msm} to determine the most likely
underlying states for all patients. The results for patient 100103 are
shown, who appeared to `recover' to a less severe state of disease
while in state 3.  We assume this is not biologically possible for the
true states, so we expect that either the observation of state 3 at
time 4.98 was an erroneous observation of state 2, or their apparent
state 2 at time 5.94 was actually state 3.  According to the expected
path constructed using the Viterbi algorithm, it is the observation at
time 5.94 which is most probably misclassified.

<<>>=
vit <- viterbi.msm(cavmisc.msm)
vit[vit$subject==100103,]
@

\subsection{Fitting general hidden Markov models with \Rpackage{msm}}
\label{sec:fitting:hmm:general}

The \Rpackage{msm} package provides a framework for fitting
continuous-time hidden Markov models with general, continuous
outcomes.  As before, we use the \Rfunction{msm} function itself.

\paragraph{Specifying the hidden Markov model}

A hidden Markov model consists of two related components:
\begin{itemize}
\item the model for the evolution of the underlying Markov chain,
\item the set of models for the observed data conditionally on each
  underlying state.
\end{itemize}

The model for the transitions between underlying states is specified
as before, by supplying a \Rfunarg{qmatrix}.  The model for the
outcomes is specified using the argument \Rfunarg{hmodel} to
\Rfunction{msm}.  This is a list, with one element for each underlying
state, in order.  Each element of the list should be an object
returned by a hidden Markov model \emph{constructor function}.  The
HMM constructor functions provided with \Rpackage{msm} are listed in
Table \ref{tab:hmm:dists}.  There is a separate constructor function
for each class of outcome distribution, such as uniform, normal or
gamma.

Consider a three-state hidden Markov model, with a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llll}
    -q_{12} & q_{12} &  0  \\
    0  & -q_{23} & q_{23}\\
    0    &   0    &   0   \\
  \end{array}
\right )
\]
Suppose the outcome distribution for state 1 is Normal$(\mu_1,
\sigma^2_1)$, the distribution for state 2 is Normal$(\mu_2,
\sigma^2_2)$, and state 3 is exactly observed. Observations of state 3
are given a label of -9 in the data.  Here our \Rfunarg{hmodel}
argument should be a list of objects returned by \Rfunction{hmmNorm} and
\Rfunction{hmmIdent} constructor functions.

We must specify initial values for the parameters as the arguments to
the constructor functions.  For example, we take initial values of
$\mu_1 = 90, \sigma_1 = 8, \mu_2 = 70, \sigma_2 = 8$.  Initial values
for $q_{12}$ and $q_{23}$ are 0.25 and 0.2. Finally suppose the
observed data are in a variable called \Robject{y}, the measurement
times are in \Robject{time}, and subject identifiers are in
\Robject{ptnum}.  The call to \Rfunction{msm} to estimate the
parameters of this hidden Markov model would then be
\begin{Scode}
  msm( y ~ time, subject=ptnum, data = example.df,
  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
                 hmmIdent(-9)) )
\end{Scode}

\begin{table}[htbp]
\scriptsize
  \centering
  \begin{tabular}{lp{0.8in}p{0.6in}p{0.6in}p{0.7in}l}
    \hline
    Function & Distribution & Parameters &  &  Location (link) & Density for an observation $x$  \\
    \hline
    \Rfunction{hmmCat}   & Categorical   & \Rfunarg{prob, basecat}  & $p,c_0$ & $p$ (logit) & $p_x$, $x = 1, \ldots, n$  \\
    \Rfunction{hmmIdent}   & Identity  & \Rfunarg{x}  & $x_0$  & & $I_{x = x_0}$  \\
    \Rfunction{hmmUnif}   & Uniform  & \Rfunarg{lower, upper}  & $l,u$  & & $1 / (u - l)$, $u \leq x \leq l$  \\
    \Rfunction{hmmNorm}   & Normal  & \Rfunarg{mean, sd}  & $\mu,\sigma$  & $\mu$ (identity) & $\phi(x, \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp(-(x - \mu)^2/(2 \sigma^2) )$  \\
    \Rfunction{hmmLNorm}   & Log-normal  & \Rfunarg{meanlog, sdlog}  & $\mu,\sigma$  & $\mu$ (identity) & $\frac{1}{x \sqrt{2 \pi \sigma^2}} \exp(-(\log x - \mu)^2 / (2 \sigma^2))$   \\
    \Rfunction{hmmExp}   & Exponential  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda e^{- \lambda x}$, $x > 0$  \\
    \Rfunction{hmmGamma}   & Gamma  & \Rfunarg{shape, rate}  & $n,\lambda$  &  $\lambda$ (log) & $\frac{\lambda^n}{\Gamma(n)}x^{n-1} \exp(-\lambda x)$, $x > 0, n > 0, \lambda > 0$   \\
    \Rfunction{hmmWeibull}   & Weibull  & \Rfunarg{shape, scale}  & $a, b$  & $b$ (log) & $\frac{a}{b} (\frac{x}{b})^{a-1} \exp{(-(\frac{x}{b})^a)}$, $x > 0$  \\
    \Rfunction{hmmPois}   & Poisson  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda^x \exp(-\lambda)/x!$, $x = 0, 1, 2, \ldots$  \\
    \Rfunction{hmmBinom}   & Binomial  & \Rfunarg{size, prob}  & $n,p$  & $p$ (logit) & ${n \choose x} p^x (1-p)^{n-x}$  \\
    \Rfunction{hmmNBinom}   & Negative binomial  & \Rfunarg{disp, prob}  & $n,p$  & $p$ (logit) & $\Gamma(x+n)/(\Gamma(n)x!) p^n (1-p)^x$ \\
    \Rfunction{hmmBeta}   & Beta  & \Rfunarg{shape1,shape2}  & $a,b$  & & $ \Gamma(a+b) / (\Gamma(a)\Gamma(b))x^{a-1}(1-x)^{b-1}$ \\
    \Rfunction{hmmT}   & Student $t$  & \Rfunarg{mean, scale, df}  & $\mu,\sigma,k$  &  $\mu$ (identity) & $\frac{\Gamma\left((k+1)/2\right)}{\Gamma(k/2)}{\sqrt{\frac{1}{k\pi\sigma^2}}}\left\{1 + \frac{1}{k\sigma^2}(x - \mu)^{2} \right\}^{-(k + 1)/2}$ \\
    \Rfunction{hmmTNorm}   & Truncated normal  & \Rfunarg{mean, sd, lower, upper}  & $\mu,\sigma,l,u$  & $\mu$ (identity) &
\parbox[t]{2in}{$\phi(x, \mu, \sigma) / \\ (\Phi(u, \mu, \sigma) - \Phi(l, \mu, \sigma))$, \\ where $\Phi(x,\mu,\sigma) = \int_{-\infty}^x \phi(u,\mu,\sigma) du$}
 \\
    \Rfunction{hmmMETNorm}   & Normal with truncation and measurement error  & \Rfunarg{mean, sd, lower, upper, sderr, meanerr}  & \parbox[t]{1in} {$\mu_0,\sigma_0,l,u$, \\ $\sigma_\epsilon,\mu_\epsilon$}  &  $\mu_\epsilon$ (identity) &
\parbox[t]{2in}{$( \Phi(u, \mu_2, \sigma_3) - \Phi(l, \mu_2, \sigma_3)) / $ \\
$(\Phi(u, \mu_0, \sigma_0) - \Phi(l, \mu_0, \sigma_0)) $ \\
$\times \phi(x, \mu_0 + \mu_\epsilon, \sigma_2)$, \\
$\sigma_2^2 = \sigma_0^2 + \sigma_\epsilon^2$, \\
$\sigma_3 = \sigma_0 \sigma_\epsilon / \sigma_2$, \\
$\mu_2 = (x - \mu_\epsilon) \sigma_0^2 + \mu_0 \sigma_\epsilon^2$}
 \\
    \Rfunction{hmmMEUnif}   & Uniform with measurement error  & \Rfunarg{lower, upper, sderr, meanerr}  & $l,u,\mu_\epsilon,\sigma_\epsilon$  & $\mu_\epsilon$ (identity) &
    \parbox[t]{2in}{$(\Phi(x, \mu_\epsilon+l, \sigma_\epsilon) - \Phi(x, \mu_\epsilon+u, \sigma_\epsilon)) / \\ (u - l)$} \\
    \hline
  \end{tabular}
  \caption{Hidden Markov model distributions in \Rpackage{msm}.}
  \label{tab:hmm:dists}
\end{table}

\paragraph{Covariates on hidden Markov model parameters}

Most of the outcome distributions can be parameterised by covariates,
using a link-transformed linear model.  For example, an observation
$y_{ij}$ may have distribution $f_1$ conditionally on underlying state
1. The link-transformed parameter $\theta_1$ is a linear function of
the covariate vector $x_{ij}$ at the same observation time.
\begin{eqnarray*}
  \label{eq:hmm:covs}
  y_{ij} | S_{ij} &  \sim &   f_1 (y | \theta_1, \gamma_1)\\
  g(\theta_1) & = & \alpha + \beta^T x_{ij}
\end{eqnarray*}
Specifically, parameters named as the ``Location'' parameter in
Table \ref{tab:hmm:dists} can be modelled in terms of covariates, with
the given link function.

The \Rfunarg{hcovariates} argument to \Rfunction{msm} specifies the
model for covariates on the hidden Markov outcome distributions.  This
is a list of the same length as the number of underlying states, and
the same length as the \Rfunarg{hmodel} list.  Each element of the
list is a formula, in standard R linear model notation, defining the
covariates on the distribution for the corresponding state.  If there
are no covariates for a certain hidden state, then insert a \texttt{NULL} in
the corresponding place in the list.  For example, in the three-state
normal-outcome example above, suppose that the normal means on states
1 and 2 are parameterised by a single covariate $x$.
\[
\mu_1 = \alpha_1 + \beta_1 x_{ij}, \qquad \mu_2 = \alpha_2 + \beta_2 x_{ij}.
\]
The equivalent call to \Rfunction{msm} would be
\begin{Scode}
  msm( state ~ time, subject=ptnum, data = example.df,
  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
                 hmmIdent(-9)),
  hcovariates = list ( ~ x, ~ x, NULL)
 ).
\end{Scode}

\paragraph{Constraints on hidden Markov model parameters}

Sometimes it is realistic that parameters are shared between some of
the state-specific outcome distributions.  For example, the
Normally-distributed outcome in the previous example could have a
common variance $\sigma^2_1 = \sigma^2_2 = \sigma^2$ between states 1
and 2, but differing means.  It would also be realistic for any
covariates on the mean to have a common effect $\beta_1 = \beta_2 =
\beta$ on the state 1 and 2 outcome distributions.

The argument \Rfunarg{hconstraint} to \Rfunction{msm} specifies which
hidden Markov model parameters are constrained to be equal.  This is a
named list.  Each element is a vector of constraints on the named
hidden Markov model parameter. The vector has length equal to the
number of times that class of parameter appears in the whole model.
As for the other constraint arguments such as \Rfunarg{qconstraint},
identical values of this vector indicate parameters constrained to be
equal.

For example, consider the three-state hidden Markov model described
above, with normally-distributed outcomes for states 1 and 2.
To constrain the outcome variance to be equal for states 1 and 2,
and to also constrain the effect of \Robject{x} on the outcome mean
to be equal for states 1 and 2, specify
\begin{Scode}
hconstraint = list(sd = c(1,1), x=c(1,1))
\end{Scode}

Parameters of the
outcome distributions may also be constrained within specific ranges.
If chosen carefully, this may improve identifiability of hidden Markov states.
For example to constrain the mean for state 1 to be between 80 and 110,
and the mean for state 2 to be between 50 and 80, specify
\begin{Scode}
hranges = list(mean=list(lower=c(80,50), upper=c(110,80)))
\end{Scode}
Maximum likelihood estimation is then performed on the appropriate log
or logit-transformed scale so that these constraints are
satisfied. See the \Rfunction{msm} help page for further details.
Note that initial values should be strictly within the ranges, and not
on the range boundary.

\paragraph{FEV$_1$ after lung transplants}

Now we give an example of fitting a hidden Markov model to a real
dataset.  The data on FEV$_1$ measurements from lung transplant
recipients, described in \ref{sec:hmm:example:fev}, are provided with
the \Rpackage{msm} package in a dataset called \Robject{fev}.
We fit models Models 1 and 2, each with three states and common $Q$
matrix.
<<>>=
three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
@

The simpler Model 1 is specified as follows. Under this model the
FEV$_1$ outcome is Normal with unknown mean and variance, and the mean
and variance are different between BOS state 1 and state 2.
\Rfunarg{hcovariates} specifies that the mean of the Normal outcome
depends linearly on acute events.  Specifically, this covariate is an
indicator for the occurrence of an acute event within 14 days of the
observation, denoted \texttt{acute} in the data.  As an initial guess,
we suppose the mean FEV$_1$ is 100\% baseline in state 1, and 54\%
baseline in state 2, with corresponding standard deviations 16 and 18,
and FEV$_1$ observations coinciding with acute events are on average
8\% baseline lower. \Rfunarg{hconstraint} specifies that the acute
event effect is equal between state 1 and state 2.

Days of death are coded as 999 in the \texttt{fev} outcome variable.

<<>>=
hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18),
                hmmIdent(999))

fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
                death=3, hmodel=hmodel1,
                hcovariates=list(~acute, ~acute, NULL),
                hcovinits = list(-8, -8, NULL),
                hconstraint = list(acute = c(1,1)), method="BFGS")

fev1.msm

sojourn.msm(fev1.msm)
@
